Accurate binding of sodium and calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-11-16 Manuscript/manuscript.tex

In this work, we improve the cation binding of a POPC bilayer in MD simulations via the implicit inclusion of the electronic polarizability in the polar region of phospholipids. For this, we apply a simple mean-field electronic continuum correction (ECC)~\cite{leontyev11} which implies scaling the partial charges of the atoms to account for the missing electronic polarizability in MD simulations. Such an approach has been previously shown to improve the behavior of ions in bulk water simulations ~\cite{martinek17, Pluharova2014, kohagen14, kohagen16}. As a starting template for our POPC model, we will use the Lipid14~\cite{dickson14}. This force field provides one of the best available descriptions of cation binding \cite{catte16}. The newly developed ECC-lipids model reproduces the experimentally
measurable structural parameters of an ion-free POPC lipid bilayer with the accuracy comparable to the best state of the art lipid models, while improving the membrane binding affinities significantly to sodium and calcium cations.

@@

\subsection{POPC membrane structure and dynamics}

\begin{figure}[tb!]
  \centering
  \includegraphics[width=8.2cm]{../Fig/ipython_nb/Order-parameters_form-factors_exp-L14-ECCL17_q80_sig89.pdf}
  \includegraphics[width=7.4cm]{../Fig/POPCstructure.eps}
  %\includegraphics[width=8.0cm]{../Fig/ipython_nb/Order-parameters_exp-L14-ECCL17_q80_sig89.pdf}
  %\includegraphics[height=8.4cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
  \caption{\label{simVSexpNOions}
    Top: X-ray scattering form factors from simulations with the Lipid14 \cite{dickson14} and
    the ECC-lipids models compared with experiments~\cite{kucerka11} at {\color{red}??K}.
    Middle: Order parameters of POPC head group, glycerol backbone and acyl chains 
    from simulations with the Lipid14 \cite{dickson14} and the ECC-lipids models
    compared with experiments \cite{ferreira13} at {\color{red}??K}.
    The size of the markers for the head group order parameters correspond to
    the error estimate $\pm 0.02$ for experiments \cite{botan15,ollila16},
    while the error estimate for simulations is $\pm 0.005$.
    The size of the points for acyl chains are decreased by a factor of 3 to improve the clarity of the plot.
    Bottom: The chemical structure of POPC and the labeling of the carbon segments.
    {\color{red}Joe, please add the temperatures}
  } 
\end{figure}

\begin{table}[tb!]
  \caption{Values of the area per lipid (APL) of POPC bilayers without ions from the Lipid14 simulation
    ran in this work and from the literature \cite{dickson14}, from the ECC-lipids model, and from experiments. \label{tab:apls}
  }
    %    { \color{red} HECTOR:Are we using here the same water model? If not is it fair the comparison considering the large changes observed in the SI table?
    %      SAMULI: For Lipid14 simulations the water is TIP3P and SPC/E for ECC-lipids. I do understand why the comparison would
    %    not be fair in this context?}
  \begin{tabular}{l|c c}
    model          & APL (\AA$^2$)   & Temperature [K] \\
    \hline
    Lipid14                   & 65.1$\pm$ 0.6  &  300 \\
    Lipid14 \cite{dickson14}  & 65.6$\pm$ 0.5  &  303 \\
    \hline
    ECC-lipids                & 63.2$\pm$ 0.6  &  300       \\
    \hline
    experiment \cite{kucerka11} & 64.3  &  303    \\
    \hline
  \end{tabular}
\end{table}

First, we present results for bilayers in pure water.
The ECC-lipids and Lipid14 models both reproduce the experimental x-ray scattering form factors
of a POPC bilayer with a comparable accuracy in Fig.~\ref{simVSexpNOions}.
The area per lipid from the Lipid14 model is $\approx$1\AA{} larger than the
experimental value in Table~\ref{tab:apls}, while the value from the ECC-lipids model
is $\approx$1\AA{} smaller
%\todo{HECTOR: The 63.2 value does not appear in the SI table. There it is 62.2 but I'm not sure corresponds to the same system
%SAMULI: I can find the 63.2 from the SI table with SPE/E and the value 62.2 with OPC3, so everything seems to be ok for me.}.
The values of the area per lipid of the ECC-lipids model show larger variations
when simulated with different water models (i.e., 62.2--66.8 \AA{}, see Table~S2 in SI,
while still being close to the experimentally reported values.
We can thus conclude that the ECC-lipids model reproduces the experimental dimensions of the POPC
lipid bilayer with the comparable accuracy to other state of the art lipid models~\cite{ollila16}.
%The optimal value for $f_\sigma$ was found by representing the overall membrane structure well
%by matching scattering form factors to experimental data~\cite{Petrache06, Kucerka08, Pabst10}.

Similarly, the acyl chain order parameters of the ECC-lipids model, as well as the Lipid14 model~\cite{dickson14}, agree with the experimental values within the error bars, as presented in Fig.~\ref{simVSexpNOions}. Notably, the experimentally measured forking and small order parameter values of $C_2$ segment in {\it sn}-2 chain are well reproduced by both models. This feature has been suggested to indicate that the carbonyl of {\it sn}-2 chain is directed towards the water phase, in contrast to the carbonyl in {\it sn}-1 chain, which would orient more along the bilayer plane~\cite{seelig75,schindler75,gawrisch92}.
While this arrangement may be a relevant feature for the ion binding details, it is not fully reproduced by other available lipid models~\cite{ollila16}.
%This is expected as the ECC model does not modify the already highly optimized parameters of this region in Lipid14

The order parameters of the $\alpha$ and $\beta$ carbons in the headgroup are slightly larger in the ECC-lipids model than in the Lipid14 model, which is apparently related to the P-N vector orienting by about 7$^{\circ}$ more toward the water phase in the ECC-lipids model, see Fig.~\ref{OrderParameterCHANGESsurf}. Considering the available experimental evidence, it is not possible to suggest which of the two models provides more realistic headgroup orientation. The ECC-lipids model gives the $\beta$ carbon order parameter value closer to experiments than the Lipid14 model, while the opposite is true for the $\alpha$ carbon. The accuracy of both models in the glycerol backbone region is equally footed with any other state of art lipid model available in literature \cite{botan15}, see Fig.~\ref{simVSexpNOions}.

@@

\subsection{Calibration of lipid electrometer:
            Response of POPC head groups to bound charge}\label{section:boundCHARGE}

@@

The changes of the headgroup order parameters with increasing amount of the cationic surfactant are compared between simulations and experiments~\cite{scherer89} in Fig.~\ref{OrderParameterCHANGESsurf}. An approximately linear decrease of the order parameters, as expected from Eq. \ref{OPchangeEQ}, is observed in both simulations and experiments at least for mole fractions below $\sim$30\%. The slope is, however, too steep in the Lipid14 model indicating that the response of head group order parameters is too large to the bound positive charge. In contrast, the slope of the ECC-lipids model is in a very good agreement with experiments for the $\alpha$ segment, while being slightly underestimated for the $\beta$ segment.

@@

\subsection{Binding affinities to POPC membrane validated through lipid electrometer}

@@

The results from the simulations combining the ECC-lipids with the ECC-ion models \cite{martinek17, kohagen16, Pluharova2014} exhibit a significantly improved behavior of the POPC head group order parameters as functions of NaCl or CaCl$_2$ concentrations, see Fig.~\ref{fig:delta_ordPar_NaCl} and Fig.~\ref{fig:delta_ordPar_CaCl}. Considering that we can also reproduce the experimental response in systems with known charge density (see above section \ref{section:boundCHARGE}), we conclude that our ECC-model correctly reproduces the binding affinities of Na$^{+}$ and Ca$^{2+}$ ions to the POPC lipid bilayer. Furthermore, while the response of the glycerol backbone $g_3$ order parameter to CaCl$_2$ was significantly overestimated in the original Lipid14 model, the ECC-lipids model provides an improved agreement with experiment, as seen in Fig.~\ref{fig:delta_ordPar_CaCl}.
%\todo{HECTOR: I do not believe in the content of the following sentence as it is. How do we know that the changes of the P-N angle are too pronounced for Lipid14?. There is no direct correlation between angle and order parameter.}
% SAMULI: The sentence is now modified.
Also the changes of the P-N vector angle are more pronounced for the Lipid14 model,
for which the largest tilting toward water phase induced by a $780\,\mathrm{mM}$
CaCl$_2$ concentration is approximately 17$^{\circ}$. The corresponding value
for the ECC-POPC simulation is only 6$^{\circ}$ ($820\,\mathrm{mM}$ CaCl$_2$). 
%\todo{HECTOR: I have a problem when comparing fig. 2 and 3. If I read it right the behavior when adding calcium and surfactant is not the same. Is the latter changing the area per lipid and therefore the surface charge density? If this will be the case then besides the effect of the charge we will have a packing effect. This is more relevant for the more concentrated cases. Shouldn't we try to discuss the two figures and what the differences and similarities mean? Given our previous work, should we check the behavior respect the charge density?}
%SAMULI: Not now, hopefully in the future.

@@

Finally, the ion binding affinities for the ECC-lipids model with different water models are compared in SI. In general, the performance of ECC-lipids with any of the water models is better than that of the original Lipid14 model, with the order parameter changes being slightly overestimated with the four-site water models and with TIP3p model.

\subsection{Structure and affinity of cation binding to POPC membrane}
\label{sec:affinity}

\todo{HECTOR: I'm not sure I understand the meaning of "structure" in the title. In general the titles in the result section are much alike between them, specially the first ones. This makes difficult to guess where to find a particular piece of information. I will make an effort to make them more clear.
SAMULI: I agree with the titles, especially the word ''structure'' here. Can you suggest new ones?}

\begin{table}[tb!]
  \caption{Relative surface excess of calcium with respect to water ($\Gamma_{Ca}^{\rm water}$) in different POPC bilayer models. All systems contain initially 350~mM of Ca$^{2+}$ (C$_w$). C$_b$ correspond to the end water bulk concentration of Ca$^{2+}$. $r^\mathrm{Ca^{2+}} _\mathrm{PO_4} $ and $r^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}}$ show the location for bound calcium in percentages for phosphate and carbonyl oxygens. {\color{red}HECTOR:PLEASE CHECK}
  \label{tab:binding}}
  \begin{tabular}{l|c | c c | c c}
    model                 & $\Gamma_{Ca}^{\rm water} \,/\, \mathrm{nm}^{-2}$ & $C_{w}$ & $C_b\,/\,\mathrm{mM}$   & $r^\mathrm{Ca^{2+}} _\mathrm{PO_4} $ & $r^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}} $ \\
    \hline
    ECC-lipids            &  $0.06 \pm 0.01 $ &  350  &  $280\pm 10 $  &  99\%  &    25\%    \\
    Lipid14/\AA{}qvist    &  $0.13 \pm 0.01 $ &  350  &  $210\pm 10 $  & 100\%  &    37\%     \\
    Lipid14/Dang          &  $0.23 \pm 0.03 $ &  350  &  $160\pm 10 $  & 100\%  &    14\%    \\
    Lipid14/ECC-ions      &  $0.35 \pm 0.11 $ &  350  &  $120\pm 100$  & 100\%  &    23\%    \\
%Average no. of bound cations:
%L14+Aqvist: all:22.7, phos:22.7, carb:8.5
%L14+Dang:   all:29.2, phos:29.2, carb:4.0
%L14+ECC-ion:all:37.7, phos:37.7, carb:8.5
  \end{tabular}
\end{table}

Binding affinities of Ca$^{2+}$ ions to a POPC bilayer in different simulation models were quantified by calculating the relative surface excess of calcium respect to water molecules, $\Gamma_{\rm ion}^{\rm water}$, from Eq.~\ref{surfexcess}. The values of $\Gamma_{\rm ion}^{\rm water}$, shown in Table~\ref{tab:binding}, were calculated from simulations with the same initial concentration of CaCl$_2$ of 350~mM before solvating the lipids. As expected from the changes of the lipid headgroup order parameters in Fig.~ \ref{fig:delta_ordPar_CaCl}, the relative surface excess for the ECC-lipids model, $\Gamma_{\rm Ca}^{\rm water}$ = 0.06~nm$^{-2}$, is significantly smaller than for the other models, 0.13--0.35~nm$^{-2}$.
\todo{HECTOR: I have deleted the sentence relating the uncertainty and the error as in my opinion there are other factors which are more important. The most important the slow convergence. SAMULI: I think that this was a good idea.}.
Interestingly, the calculated relative surface excess of \ce{NaCl} at $1\,\mathrm{M}$ concentration (ECC-ions~\cite{Pluharova2014}) using our POPC-ECC model is not only quantitatively but also qualitatively different from \ce{CaCl2} having a negative value $\Gamma_{Na}^{water} = (-0.11 \pm 0.01) \rm{nm}^{-2}$ (Fig.~\ref{fig:cacl-dens}). This value means that water molecules are preferred to sodium and chloride ions at the membrane-water interface.  This is in contradiction with most of the available lipid force fields, which predict a positive binding of sodium to PC lipid bilayers \cite{catte16}.
%\todo{HECTOR: do we need to add a NaCl table or an extended table for CaCl2 using different concentrations in the SI? I think this is very important information
%  SAMULI: I agree that this is important information, but we have quite a lot of stuff in the paper already. Maybe we could leave this for the future studies.}

\begin{figure}[htbp!]
  \centering
  \includegraphics[height=19.5cm]{../Fig/ipython_nb/density_profiles_ca_cl_wat_phos_models-compar.pdf}
  \caption{\label{fig:cacl-dens}
    Number density profiles of \ce{Ca^{2+}}, \ce{Na^{+}} and \ce{Cl^-} along membrane normal axis
    for different force fields. Data for Lipid14 with \AA{}qvist ions are taken directly from Ref. \citenum{catte16}.
    %Normalization factor for ions is 1 for monovalent ions (i.e.~\ce{Na^+} and~\ce{Cl^-}),
    In order to visualize the density profiles with a scale comparable to the profile of \ce{Ca^{2+}}, 
    densities of~\ce{Cl^-} and \ce{Na^{+}} are divided by 2, and
    densities of phosphate group resp. water are divided by 5 resp. 200. 
    The molar concentration of ions in water is 350~mM in all systems
    presented here. 
    }
\end{figure}

We analyzed the ratio of the number of calcium cations bound to either phosphate or carbonyl moieties and the total number of bound cations in our POPC bilayers as done previously in Ref.~\citenum{javanainen17}. A maximum distance of 0.3~nm from any lipid oxygen is used to determine a bound calcium. The results in Table~\ref{tab:binding} show that almost all (99\%) of the bound Ca$^{2+}$ ions are in contact with the phosphate oxygens. From those only one third (32\%) also interact with the carbonyl oxygens. The interaction between calcium ions with only carbonyl oxygens is rare (1\%). Neither individual lipids nor source chain of carbonyl is distinguished in the analysis. The most likely interactions between Ca$^{2+}$ ions and phosphate oxygens are visualized with the probability density isocontours in Fig.~\ref{fig:volmaps}. While higher concentrations of \ce{CaCl2} increase the number of contacts per lipid, the distribution of contacts between phosphate and carbonyl oxygens is not affected.

\begin{figure}[tb!]
  \centering
  \includegraphics[width=8.0cm]{../Fig/volmap_resid10_Ca_Cl_PO4Cent.png} %\\
%  \includegraphics[width=4.0cm]{../Fig/volmap_resid10_Ca_Cl_sn1Cent.png}
%  \includegraphics[width=3.6cm]{../Fig/volmap_resid10_Ca_Cl_sn2Cent.png}
  \caption{\label{fig:volmaps}
      Isocontours of probability density of \ce{Ca^{2+}} with respect to 
      the phosphate oxygens of POPC from ECC-lipids simulation.
      The probability density was evaluated around a single lipid, 
      after its strucutral alignment using only phosphate group. 
      %moiety, side chain 1 carbonyl group and side chain 2 carbonyl group.
      %Shown contours suggest that the dominant contribution 
      %to \ce{Ca^{2+}} binding comes from the phosphate oxygens,
      %whereas the interactions with any of the two carbonyl groups are considerably milder.
  }
  \todo{JOE: I'll update this figure with some ensemble of configuration to support binding preference of \ce{Ca^{2+}}
  SAMULI: I am not sure if this would be needed anymore.}
\end{figure}

@@

Simple binding models have been used previously to interpret the experimental data \cite{altenbach84,macdonald87} we use in this work to validate the simulation models (Fig.~\ref{fig:delta_ordPar_CaCl}). In particular, NMR data about PC headgroup order parameters response and atomic absorption spectroscopy were best explained using a ternary complex binding model with a binding stoichiometry of one \ce{Ca^{2+}} per two POPC lipids~\cite{altenbach84}. However, a Langmuir adsorption model assuming a \ce{Ca^{2+}}:POPC stoichiometry of 1:1 also provided a good fit to the experimental data when considering only \ce{CaCl2}low concentrations~\cite{macdonald87}.
%\todo{HECTOR: Why is the following sentence remotely significant. I will vote for removing it} Note that the results based on these binding models and the measured headgroup order parameter changes are also included in the tabulated experimental binding constants~\cite{marsh13}.

>>>>

In this work, we improve the cation binding of a POPC bilayer in MD simulations via the implicit inclusion of the electronic polarizability in the polar region of phospholipids. For this, we apply a simple mean-field electronic continuum correction (ECC)~\cite{leontyev11} which implies scaling the partial charges of the atoms to account for the missing electronic polarizability in MD simulations. Such an approach has been previously shown to improve the behavior of ions in bulk water simulations ~\cite{martinek17, Pluharova2014, kohagen14, kohagen16}. As a starting template for our POPC model, we will use the Lipid14~\cite{dickson14}. This force field provides one of the best available descriptions of cation binding \cite{catte16}. The newly developed ECC-POPC model reproduces the experimentally
measurable structural parameters of an ion-free POPC lipid bilayer with the accuracy comparable to the best state of the art lipid models, while improving the membrane binding affinities significantly to sodium and calcium cations.

@@

\subsection{Structural parameters of pure ECC-POPC model membrane are in agreement with experiments}

\begin{figure}[tb!]
  \centering
  \includegraphics[width=8.2cm]{../Fig/ipython_nb/Order-parameters_form-factors_exp-L14-ECCL17_q80_sig89.pdf}
  \includegraphics[width=7.4cm]{../Fig/POPCstructure.eps}
  %\includegraphics[width=8.0cm]{../Fig/ipython_nb/Order-parameters_exp-L14-ECCL17_q80_sig89.pdf}
  %\includegraphics[height=8.4cm,angle=-90]{../Fig/form-f_exp-l14-eccl17.eps}
  \caption{\label{simVSexpNOions}
    Top: X-ray scattering form factors from simulations with the Lipid14 \cite{dickson14} and
    the ECC-POPC models compared with experiments~\cite{kucerka11} at {\color{red}??K}.
    Middle: Order parameters of POPC head group, glycerol backbone and acyl chains 
    from simulations with the Lipid14 \cite{dickson14} and the ECC-POPC models
    compared with experiments \cite{ferreira13} at {\color{red}??K}.
    The size of the markers for the head group order parameters correspond to
    the error estimate $\pm 0.02$ for experiments \cite{botan15,ollila16},
    while the error estimate for simulations is $\pm 0.005$.
    The size of the points for acyl chains are decreased by a factor of 3 to improve the clarity of the plot.
    Bottom: The chemical structure of POPC and the labeling of the carbon segments.
    {\color{red}Joe, please add the temperatures}
  } 
\end{figure}

\begin{table}[tb!]
  \caption{Values of the area per lipid (APL) of POPC bilayers without ions from the Lipid14 simulation
    ran in this work and from the literature \cite{dickson14}, from the ECC-POPC model, and from experiments. \label{tab:apls}
  }
    %    { \color{red} HECTOR:Are we using here the same water model? If not is it fair the comparison considering the large changes observed in the SI table?
    %      SAMULI: For Lipid14 simulations the water is TIP3P and SPC/E for ECC-lipids. I do understand why the comparison would
    %    not be fair in this context?}
  \begin{tabular}{l|c c}
    model          & APL (\AA$^2$)   & Temperature [K] \\
    \hline
    Lipid14                   & 65.1$\pm$ 0.6  &  300 \\
    Lipid14 \cite{dickson14}  & 65.6$\pm$ 0.5  &  303 \\
    \hline
    ECC-POPC                & 63.2$\pm$ 0.6  &  300       \\
    \hline
    experiment \cite{kucerka11} & 64.3  &  303    \\
    \hline
  \end{tabular}
\end{table}

First, we present results for bilayers in pure water.
The ECC-POPC and Lipid14 models both reproduce the experimental x-ray scattering form factors
of a POPC bilayer with a comparable accuracy in Fig.~\ref{simVSexpNOions}.
The area per lipid from the Lipid14 model is $\approx$1\AA{} larger than the
experimental value in Table~\ref{tab:apls}, while the value from the ECC-POPC model
is $\approx$1\AA{} smaller
%\todo{HECTOR: The 63.2 value does not appear in the SI table. There it is 62.2 but I'm not sure corresponds to the same system
%SAMULI: I can find the 63.2 from the SI table with SPE/E and the value 62.2 with OPC3, so everything seems to be ok for me.}.
The values of the area per lipid of the ECC-POPC model show larger variations
when simulated with different water models (i.e., 62.2--66.8 \AA{}, see Table~S2 in SI),
while still being close to the experimentally reported values.
We can thus conclude that the ECC-POPC model reproduces the experimental dimensions of the POPC
lipid bilayer with the comparable accuracy to other state of the art lipid models~\cite{ollila16}.
%The optimal value for $f_\sigma$ was found by representing the overall membrane structure well
%by matching scattering form factors to experimental data~\cite{Petrache06, Kucerka08, Pabst10}.

Similarly, the acyl chain order parameters of the ECC-POPC model, as well as the Lipid14 model~\cite{dickson14}, agree with the experimental values within the error bars, as presented in Fig.~\ref{simVSexpNOions}. Notably, the experimentally measured forking and small order parameter values of $C_2$ segment in {\it sn}-2 chain are well reproduced by both models. This feature has been suggested to indicate that the carbonyl of {\it sn}-2 chain is directed towards the water phase, in contrast to the carbonyl in {\it sn}-1 chain, which would orient more along the bilayer plane~\cite{seelig75,schindler75,gawrisch92}.
While this arrangement may be a relevant feature for the ion binding details, it is not fully reproduced by other available lipid models~\cite{ollila16}.
%This is expected as the ECC model does not modify the already highly optimized parameters of this region in Lipid14

The order parameters of the $\alpha$ and $\beta$ carbons in the headgroup are slightly larger in the ECC-POPC model than in the Lipid14 model, which is apparently related to the P-N vector orienting by about 7$^{\circ}$ more toward the water phase in the ECC-POPC model, see Fig.~\ref{OrderParameterCHANGESsurf}. Considering the available experimental evidence, it is not possible to suggest which of the two models provides more realistic headgroup orientation. The ECC-POPC model gives the $\beta$ carbon order parameter value closer to experiments than the Lipid14 model, while the opposite is true for the $\alpha$ carbon. The accuracy of both models in the glycerol backbone region is equally footed with any other state of art lipid model available in literature \cite{botan15}, see Fig.~\ref{simVSexpNOions}.

@@

\subsection{Calibration of head group response to membrane-bound charge using cationic surfactant}\label{section:boundCHARGE}

@@

The changes of the headgroup order parameters with increasing amount of the cationic surfactant are compared between simulations and experiments~\cite{scherer89} in Fig.~\ref{OrderParameterCHANGESsurf}. An approximately linear decrease of the order parameters, as expected from Eq. \ref{OPchangeEQ}, is observed in both simulations and experiments at least for mole fractions below $\sim$30\%. The slope is, however, too steep in the Lipid14 model indicating that the response of head group order parameters is too large to the bound positive charge. In contrast, the slope of the ECC-POPC model is in a very good agreement with experiments for the $\alpha$ segment, while being slightly underestimated for the $\beta$ segment.

@@

\subsection{Validation ECC-POPC model using binding affinities to Na$^+$ and Ca$^{2+}$ Cations: the electrometer concept}

@@

The results from the simulations combining the ECC-POPC with the ECC-ion models \cite{martinek17, kohagen16, Pluharova2014} exhibit a significantly improved behavior of the POPC head group order parameters as functions of NaCl or CaCl$_2$ concentrations, see Fig.~\ref{fig:delta_ordPar_NaCl} and Fig.~\ref{fig:delta_ordPar_CaCl}. Considering that we can also reproduce the experimental response in systems with known charge density (see above section \ref{section:boundCHARGE}), we conclude that our ECC-model correctly reproduces the binding affinities of Na$^{+}$ and Ca$^{2+}$ ions to the POPC lipid bilayer. Furthermore, while the response of the glycerol backbone $g_3$ order parameter to CaCl$_2$ was significantly overestimated in the original Lipid14 model, the ECC-POPC model provides an improved agreement with experiment, as seen in Fig.~\ref{fig:delta_ordPar_CaCl}.
%\todo{HECTOR: I do not believe in the content of the following sentence as it is. How do we know that the changes of the P-N angle are too pronounced for Lipid14?. There is no direct correlation between angle and order parameter.}
% SAMULI: The sentence is now modified.
Also the changes of the P-N vector angle are more pronounced for the Lipid14 model,
for which the largest tilting toward water phase induced by a $780\,\mathrm{mM}$
CaCl$_2$ concentration is approximately 17$^{\circ}$. The corresponding value
for the ECC-POPC simulation is only 6$^{\circ}$ ($820\,\mathrm{mM}$ CaCl$_2$). 
%\todo{HECTOR: I have a problem when comparing fig. 2 and 3. If I read it right the behavior when adding calcium and surfactant is not the same. Is the latter changing the area per lipid and therefore the surface charge density? If this will be the case then besides the effect of the charge we will have a packing effect. This is more relevant for the more concentrated cases. Shouldn't we try to discuss the two figures and what the differences and similarities mean? Given our previous work, should we check the behavior respect the charge density?}
%SAMULI: Not now, hopefully in the future.

@@

Finally, the ion binding affinities for the ECC-POPC model with different water models are compared in SI. In general, the performance of ECC-POPC with any of the water models is better than that of the original Lipid14 model, with the order parameter changes being slightly overestimated with the four-site water models and with TIP3p model.

\subsection{Binding affinity of Na$^+$ and Ca$^{2+}$ cations to POPC membrane}
\label{sec:affinity}

%\todo{HECTOR: I'm not sure I understand the meaning of "structure" in the title. In general the titles in the result section are much alike between them, specially the first ones. This makes %difficult to guess where to find a particular piece of information. I will make an effort to make them more clear.
%SAMULI: I agree. There are now new titles.}

\begin{table}[tb!]
  \caption{Bulk concentrations of Ca$^{2+}$ (C$_b$), relative surface excess of calcium with respect to water ($\Gamma_{Ca}^{\rm water}$),
    and the percetages of Ca$^{2+}$ bound to phosphate or carbonyl oxyges ($r^\mathrm{Ca^{2+}} _\mathrm{PO_4} $ and $r^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}}$)
    in different POPC bilayer models. All systems have the same molar concentration of Ca$^{2+}$ respect to water (C$_w$=350mM).
    %{\color{red}HECTOR:PLEASE CHECK}
    {\color{red}Joe, please confirm that $r^\mathrm{Ca^{2+}} _\mathrm{PO_4}$ is really 100\% for all the Lipid14 simulations
      (i.e. that this is not a typo).}
  \label{tab:binding}}
  \begin{tabular}{l|c c | c | c c}
    model                  & $C_{w}$ & $C_b\,/\,\mathrm{mM}$ & $\Gamma_{Ca}^{\rm water} \,/\, \mathrm{nm}^{-2}$  & $r^\mathrm{Ca^{2+}} _\mathrm{PO_4} $ & $r^\mathrm{Ca^{2+}} _\mathrm{O_{carb.}} $ \\
    \hline
    ECC-POPC             &  350  &  $280\pm 10 $  &  $0.06 \pm 0.01 $                           &  99\%  &    25\%    \\
    Lipid14/\AA{}qvist     &  350  &  $210\pm 10 $  &  $0.13 \pm 0.01 $                          & 100\%  &    37\%     \\
    Lipid14/Dang           &  350  &  $160\pm 10 $  &  $0.23 \pm 0.03 $                            & 100\%  &    14\%    \\
    Lipid14/ECC-ions       &  350  &  $120\pm 100$  &  $0.35 \pm 0.11 $                         & 100\%  &    23\%    \\
%Average no. of bound cations:
%L14+Aqvist: all:22.7, phos:22.7, carb:8.5
%L14+Dang:   all:29.2, phos:29.2, carb:4.0
%L14+ECC-ion:all:37.7, phos:37.7, carb:8.5
  \end{tabular}
\end{table}

Binding affinities of Ca$^{2+}$ ions to a POPC bilayer in different simulation models were quantified by calculating the relative surface excess of calcium respect to water molecules, $\Gamma_{\rm ion}^{\rm water}$, from Eq.~\ref{surfexcess}.
The values of $\Gamma_{\rm ion}^{\rm water}$
from different POPC simulations with then same molar concetration respect
to water ($C_w$=350mM) are shown in Table~\ref{tab:binding}.
As expected from the changes of the lipid headgroup order parameters in Fig.~ \ref{fig:delta_ordPar_CaCl}, the relative surface excess for the ECC-POPC model, $\Gamma_{\rm Ca}^{\rm water}$ = 0.06~nm$^{-2}$, is significantly smaller than for the other models, 0.13--0.35~nm$^{-2}$.
%\todo{HECTOR: I have deleted the sentence relating the uncertainty and the error as in my opinion there are other factors which are more important. The most important the slow convergence. SAMULI: I think that this was a good idea.}.
Interestingly, the calculated relative surface excess of \ce{NaCl} at $1\,\mathrm{M}$ concentration (ECC-ions~\cite{Pluharova2014}) using our POPC-ECC model is not only quantitatively but also qualitatively different from \ce{CaCl2} having a negative value $\Gamma_{Na}^{water} = (-0.11 \pm 0.01) \rm{nm}^{-2}$ (Fig.~\ref{fig:cacl-dens}). This value means that water molecules are preferred to sodium and chloride ions at the membrane-water interface.  This is in contradiction with most of the available lipid force fields, which predict a positive binding of sodium to PC lipid bilayers \cite{catte16}.
%\todo{HECTOR: do we need to add a NaCl table or an extended table for CaCl2 using different concentrations in the SI? I think this is very important information
%  SAMULI: I agree that this is important information, but we have quite a lot of stuff in the paper already. Maybe we could leave this for the future studies.}

\begin{figure}[htbp!]
  \centering
  \includegraphics[height=19.5cm]{../Fig/ipython_nb/density_profiles_ca_cl_wat_phos_models-compar.pdf}
  \caption{\label{fig:cacl-dens}
    Number density profiles of \ce{Ca^{2+}}, \ce{Na^{+}} and \ce{Cl^-} along membrane normal axis
    for different force fields.
    %Data for Lipid14 with \AA{}qvist ions are taken directly from Ref. \citenum{catte16}.
    %Normalization factor for ions is 1 for monovalent ions (i.e.~\ce{Na^+} and~\ce{Cl^-}),
    {\color{red} Joe, please clarify the sentence below}
    In order to visualize the density profiles with a scale comparable to the profile of \ce{Ca^{2+}}, 
    densities of~\ce{Cl^-} and \ce{Na^{+}} are divided by 2, and
    densities of phosphate group resp. water are divided by 5 resp. 200. 
    All the simulations shown here have the same molar concentration of ions in water ($C_w$=350~mM).
    }
\end{figure}

\subsection{Molecular interactions between Na$^+$ or Ca$^{2+}$ cations and POPC oxygens}
We analyzed the ratio of the number of calcium cations bound to either phosphate or carbonyl moieties and the total number of bound cations in our POPC bilayers as done previously in Ref.~\citenum{javanainen17}. A maximum distance of 0.3~nm from any lipid oxygen is used to determine a bound calcium. The results from ECC-POPC simulation in Table~\ref{tab:binding} show that almost all (99\%) of the bound Ca$^{2+}$ ions are in contact with the phosphate oxygens. From those only one third (32\%) also interact with the carbonyl oxygens. The interaction between calcium ions with only carbonyl oxygens is rare (1\%). Neither individual lipids nor source chain of carbonyl is distinguished in the analysis. The most likely interactions between Ca$^{2+}$ ions and phosphate oxygens are visualized with the probability density isocontours in Fig.~\ref{fig:volmaps}. While higher concentrations of \ce{CaCl2} increase the number of contacts per lipid, the distribution of contacts between phosphate and carbonyl oxygens is not affected.

\begin{figure}[tb!]
  \centering
  \includegraphics[width=8.0cm]{../Fig/volmap_resid10_Ca_Cl_PO4Cent.png} %\\
%  \includegraphics[width=4.0cm]{../Fig/volmap_resid10_Ca_Cl_sn1Cent.png}
%  \includegraphics[width=3.6cm]{../Fig/volmap_resid10_Ca_Cl_sn2Cent.png}
  \caption{\label{fig:volmaps}
      Isocontours of probability density of \ce{Ca^{2+}} with respect to 
      the phosphate oxygens of POPC from ECC-POPC simulation.
      The probability density was evaluated around a single lipid, 
      after its strucutral alignment using only phosphate group. 
      %moiety, side chain 1 carbonyl group and side chain 2 carbonyl group.
      %Shown contours suggest that the dominant contribution 
      %to \ce{Ca^{2+}} binding comes from the phosphate oxygens,
      %whereas the interactions with any of the two carbonyl groups are considerably milder.
  }
  \todo{The figure should be updated as discussed last week.}
\end{figure}

@@

Simple binding models have been used previously to interpret the same experimental data \cite{altenbach84,macdonald87} used in this work to validate the simulation models (Fig.~\ref{fig:delta_ordPar_CaCl}). In particular, NMR data about PC headgroup order parameters response and atomic absorption spectroscopy were best explained using a ternary complex binding model with a binding stoichiometry of one \ce{Ca^{2+}} per two POPC lipids~\cite{altenbach84}. However, a Langmuir adsorption model assuming a \ce{Ca^{2+}}:POPC stoichiometry of 1:1 also provided a good fit to the experimental data when considering only \ce{CaCl2}low concentrations~\cite{macdonald87}.
%\todo{HECTOR: Why is the following sentence remotely significant. I will vote for removing it} Note that the results based on these binding models and the measured headgroup order parameter changes are also included in the tabulated experimental binding constants~\cite{marsh13}.

In this work, we reproduce the same experimental data used to calculate binding stoichiometry with our ECC-POPC model. Using our simulations, we have direct access to exact details of the binding stoichiometry without requiring any middle model as in experiments~\cite{altenbach84, macdonald87}. To evaluate the relative propensities for each stoichiometric complex (i.e.,~1~Ca$^{2+}$:~n~POPC), we calculated for each bound Ca$^{2+}$ the number of POPC molecules within a distance of 0.3~nm.
Results from the POPC bilayer simulation with a 285~mM bulk concentration of CaCl$_2$ are shown in Fig.~\ref{fig:cacl_complexes}.
We found the largest propensity for the 1:2 complex (41\%).
However, the probabilities of complexes with the stoichiometries of 1:1~(25\%)~and 1:3~(34\%)~are only slightly lower. This variability suggests a more complex binding model than the previously considered in a simple ternary complex model. With a broad brushstroke, the simulation data can, however, be viewed such that one calcium binds to two lipids on average, because the probabilities of the complexes with 1 or 3 lipids are almost equal to each other  (and complexes with more than three lipids per one calcium ion were not observed). This probably explains why the simple the ternary complex model fits adequately the experimental data, as well as the ECC-POPC simulation results (see Fig.~S3 in SI).
